---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  warning = FALSE,
  message = FALSE
)
```

# climateR

<!-- badges: start -->
 [![DOI](https://zenodo.org/badge/158620263.svg)](https://zenodo.org/badge/latestdoi/158620263)
[![R CMD Check](https://github.com/mikejohnson51/climateR/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/mikejohnson51/climateR/actions/workflows/R-CMD-check.yaml)
[![Dependencies](https://img.shields.io/badge/dependencies-7/25-orange?style=flat)](#)
[![License: MIT](https://img.shields.io/badge/License-MIT-yellow.svg)](https://choosealicense.com/licenses/mit/)
[![Project Status: Active](https://www.repostatus.org/badges/latest/active.svg)](https://www.repostatus.org/#active)
[![codecov](https://codecov.io/gh/mikejohnson51/climateR/branch/master/graph/badge.svg?token=7zs6C91SDw)](https://codecov.io/gh/mikejohnson51/climateR)
<!-- badges: end -->

`climateR` simplifies the steps needed to get climate data into R. At its core it provides three main things:

1. A catalog of over 100,000k datasets from over 2,000 data providers/archives. See (climateR::params)

```{r}
nrow(params)
length(unique(params$id))
length(unique(params$asset))
```

This catalog is an ever evolving, set of documented and federated datasets that can be accessed by the data access utilites.

2. A general [toolkit for accessing remote and local gridded data](docs/reference/index.html#data-access) files bounded by space, time, and variable constraints (`dap`, `dap_crop`, `read_dap_file`)


3. A set of [shortcuts](docs/reference/index.html#shortcuts) that implement these methods for a core set of selected catalog elements

# Installation

```{r, eval = FALSE }
remotes::install_github("mikejohnson51/AOI") # suggested!
remotes::install_github("mikejohnson51/climateR")
```

# Basic Usage

## Rainfall from GridMet for Colorado
### October 29,1991 - November 6, 1991

```{r}
library(AOI)
library(terra)
library(climateR)

AOI = aoi_get(state = "CO", county = "all")
plot(AOI$geometry)
```

```{r}
d = getGridMET(AOI,
               varname = "pr",
               startDate = "1991-10-29",
               endDate  = "1991-11-06")

plot(d$precipitation_amount)
```


